Automatic Matching of Bullet Lands

### 2017-01-12 imaging-paper.tex
Rebuild both papers with no cache
Paper edits from Dr. Carriquiry

Each fired bullet is provided in the form of a set of six x3p files, where each file is a surface scan between adjacent grooves on the bullet, called a ``land". In the Hamby data, typical length (groove-to-groove) of a land is about 2005.9 micrometers or 2.01 millimeters. For notational simplicity, we refer to a particular land of some bullet as bullet X-Y, where X is the bullet identifier, and Y is the land number. An example of plotting one of these lands is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}. These figures show side and top profiles of the land respectively. The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artifact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clockwise or counter-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic, i.e.\ a feature that pertains to a particular class of firearms, and is not unique at the individual barrel/bullet level.

@@

\end{knitrout}
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between signatures (c).}
\end{figure}
%\afterpage{\clearpage}
Figure~\ref{fig:match} gives an overview of the automated matching routine: 
We first identify a stable region for each bullet land and extract the signature at the lowest height in this region, because typically, individual characteristics are best expressed at the lower end of the bullet (see Supplement section~\ref{supp:bulletbottom} for a more detailed discussion).

All of the other steps are done on pairs of bullet lands:
\begin{enumerate}
\item {\bf Smooth the two signatures} using a loess with a very small span (see Figure~\ref{fig:smooth}). 
\item Use cross-correlation to {\bf find the best alignment} of the two signatures: shift one of the signatures by the lag indicated by the cross-correlation function (see Figure~\ref{fig:ccf} for the cross-correlation function and Figure~\ref{fig:crosscutX} for the resulting shift).
\item Using a rolling average, {\bf identify peaks and valleys} for each of the signatures. We then define an interval around the location of the extrema on each side as one third of the distance to the location of the next extrema (see Figure~\ref{fig:smoothcutb}). Peaks and valleys constitute the \emph{striation marks} of the bullet.
\item {\bf Match striations across signatures:} based on the intervals around the extrema as defined above, we identify joint intervals between two signatures as the areas in which two or more of the individual intervals overlap: a joint interval is defined as the smallest interval that encompasses all of the overlapping intervals. A joint interval is then called a match(ing stria) between the signatures, if all of the intervals are of the same type of extrema, i.e.\ they are either all peaks or all valleys. In Figure~\ref{fig:match} all matches are shown as color-filled rectangles corresponding to their type of extrema (peaks are shown in orange, and valleys in green). Non-matching intervals are left grey. 
\item {\bf Extract features from the aligned signatures and the matches between them:} many different features could be extracted. Here, we describe a few of the ones that can be found in the literature and some that we found to be of practical relevance:
\begin{enumerate}[label=(\roman*)]
\item Maximal number of CMS (consecutive matching striae), and, similarly, the number of consecutively non-matching striae (CNMS), 
\item Number of matches and non-matches,
\item The value of the cross-correlation function (ccf) between the aligned signatures \cite{vorburger:2011},
\item Average difference $D$ between signatures, defined as the Euclidean vertical distance between surface measurements of aligned signatures. Let $f(t)$ and $g(t)$ be smoothed, aligned signatures:
\[
D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2,
\]
\item The sum $S$ of average absolute heights of matched extrema: for each of the two matched stria, compute the average of the absolute heights of the peaks or valleys. $S$ is then defined as the sum of all these averages. 
\end{enumerate}
\end{enumerate}
%
The difference $D$ between signatures is here defined as the Euclidean distance (in $\mu m$). In the paper by \citet{ma:2004}, distance is defined as a measure relative to the first signature, which serves as a comparison reference and is therefore a unitless quantity.

Counting the maximal number of CMS is part of the current practice to identify bullet matches~\citep{nichols:1997, nichols:2003, nichols:2003b}. 
In the example of Figure~\ref{fig:match}, the number of consecutive matching striations (CMS) is fifteen, a high number indicative of a match between the bullets. 
 Note that the definition of CMS given in this paper does not match the one given in \citet{thompson:2013}. There, CMS is defined only in terms of matching peaks without regarding valleys. Additionally, peaks in  \citet{thompson:2013} are  used only if they can be identified and matched `within a tolerable range' between lands. The definition given here is computationally less complex, but should yield highly correlated values, because of the requirement to only consider signatures from a stable region in the land (see Section~\ref{sec:heights} for further details on stability of regions). In the Hamby study, the definition of CMS by \citet{thompson:2013} leads to approximately half of the values of CMS defined in this paper (with a correlation coefficient between the values of the two definitions of about 0.92). 
For lead bullets, such as used in the Hamby study, \citet{biasotti:1959} considered four or more consecutive peaks (corresponding to eight or more consecutive lines in our definition) to be sufficient evidence of a match.  %We will assess the claim that eight or more consecutive lines are sufficient for a match in the next section.

@@

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are quite prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/crosscuts-1}

@@

\end{knitrout}
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
For this alignment we use the cross-correlation function to find a maximal amount of agreement between the signatures~\cite{bachrach:2002, chu:2010, vorburger:2011, thompson:2013}.
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

@@

\section{Evaluation}
In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we compute the cross-correlation of signatures extracted from heights 25$\mu m$ apart. For a stable region, we require a minimum of 0.95 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands has scratch marks across the surface, also known as `tank rash'.
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 9 Bullet 2-4}
\includegraphics[width=\textwidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet B-2}
\includegraphics[width=\textwidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet Q-4}
\includegraphics[width=\textwidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}Images of the four lands that got flagged during the quality assessment. All of them show scratch marks (tank rash) across the striation marks from the barrel. They are excluded from the remainder of the analysis.}
\end{figure}
%
We exclude these four lands from further matching considerations and
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are comparing $15 \times 6 -2 = 90 - 2 = 88$ lands from unknown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). 
 When things go well, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land C-3 and all 118 lands from known bullets. Two lands show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and~\ref{fig:hamby-perfect}c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all lands, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

@@

For smaller numbers of CMS, the percentage of false positives quickly increases. However, if we take other features of the image into account, we can increase the number of correct matches considerably: Figure~\ref{fig:densities} gives an overview of the densities of all of the features derived earlier, distinguishing between known matches (KM) and known non-matches (KNM). The densities of almost all of the features show strong differences between matches and non matches. For example, a high amount of cross-correlation between two signatures is indicative of a match --  in the Hamby study, only known matches have a cross-correlation of 0.75 or higher. There are 97 land-to-land comparisons with a cross-correlation that high.

@@

%Similarly, the Euclidean distance $D$ between two signatures is strongly correlated to known matches. Out of the 48 pairs of land-to-land signatures with a distance of less than .25, 47 correspond to known matches (the single known mismatch consists of a pair of signatures that both are very flat with peaks and valleys of less than 1.5$\mu m$). %The average distance between known matches in the Hamby study is $mean(bstats$D[bstats$match])\mu m$ with a standard deviation of $sd(bstats$D[bstats$match])\mu m$; while known mismatches have an average  distance of  $mean(bstats$D[!bstats$match])\mu m$ with a standard deviation of $sd(bstats$D[!bstats$match])\mu m$.
All of the features in Figure~\ref{fig:densities} show large, if not significant, differences between matches and non-matches. The predictive power of each one of these features is shown in the form of the Receiving Operating Characteristic (ROC) curves in Figure~\ref{fig:rocs}. The features are arranged in descending order according to the area under the curve (AUC).
The dots mark the equal error rate, i.e. the location on the ROC curve, where false positive and false negative error rates are the same. The smaller the value, the better. We see that in this instance a low EER goes hand in hand with high predictive power as measured in AUC.
The feature with the highest individual predictive power is $S$, the sum of the average heights of two signatures at peaks and valleys. The maximal number of CMS is only in the seventh position here. The overall high AUC values indicate that we can successfully employ machine learning methods to distinguish matches from non-matches.
%Table~\ref{tab:importance} shows an overview in the averages and corresponding standard deviations of the set of feature variables derived in section~\ref{sec:algorithm} from the 3D topological surface measurements. There are quite large differences in these averages, indicating

@@

\end{knitrout}
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 100 trees, the error rate of land-to-land comparisons stabilizes at 0.0039. This is a weighted average between false positive error rate of 0.0001 and an error rate of false negatives of 0.2267. This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known  matches and non-matches (see Figure~\ref{fig:tree-forest}).
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, even for the single tree the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched. % (see Figure~\ref{fig:treeresult}). 
%' \begin{figure}[hbtp]
%'   \centering
%' <<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\textwidth'>>=
%' bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
%' bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
%'   n = n(),
%'   pred = sum(pred > 0.5),
%'   forest = sum(forest > 0.5)
%' )
%' library(reshape2)
%' bm <- melt(data=bullets, id.var=c("bullet", "n"))
%' 
%' bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
%' qplot(data=bm, 100*value/n, reorder(bullet, value/n), size=I(2.5), 
%'       shape=prediction, colour=prediction, alpha=I(0.6)) + 
%'   theme_bw()  + ylab("") + 
%' #  scale_shape_manual(values=c(1,2)) +
%'   scale_x_continuous("Percent of correctly predicted land-to-land matches", 
%'                      breaks = c(0,25,50,75,100), limits=c(0,100)) +
%'   theme(legend.position = "bottom")
%' @
%' \caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. In the tree, at least \hh{40\% of all possible} lands of a bullet are  matched correctly, while the random forest achieves perfect results. Unknown bullet B is the hardest for the tree to match.}
%' \end{figure}
%' %
This makes the errors in the automated approach smaller than the human error in the Hamby study. Out of the 507 participants, who returned results, eight (out of $15 \times 507 = 7,605$) bullets were not matched conclusively, corresponding to a rate of 0.0011. 
%\hh{XXX calling this a false negative rate is problematic. FEs will be up in arms, because inconclusive doesn't necessarily mean something was missed, it could also mean that there was something wrong with the bullets. XXX }

However, error rates based on bullet-to-bullet matches do not carry a lot of weight because of the small size of the study: fifteen unknown bullets are successfully matched to two pairs of ten bullets. Matching bullets can only be tested realistically in a much bigger experiment. 
Another thing to note about  the random forest's error rates is that they are based on probability cutoffs of 0.5, i.e.\ whenever the predicted probability of a match exceeds 0.5, a match is declared. Basing this decision on 0.5 is not necessary. In practice, examiners are allowed a third option of `inconclusive'. On a probability spectrum of outcomes we could therefore introduce an interval of `inconclusive' results in the middle of the spectrum -- which turns out to be unnecessary in the Hamby study, because, here, the results from the random forest are very clear cut. Figure~\ref{fig:tree-forest} shows a comparison of the predicted probabilities of a match by the tree and the random forest. %\hh{The tree predicts 132 known matches correctly, with 40 false negatives and one false positive match. The random forest predicts all 172 known matches correctly as well as all 10,212 known non-matches.} 
%Overall, there is a lot of agreement; out of 10,800 land-to-land comparisons, only four cases result in a different decision.

@@

Besides resulting in a probabilistic quantification of matches, random forests also provide an assessment of the importance of each of the features derived from the bullets' 3D topological surface measurements. Figure~\ref{fig:importance} shows an overview of the importance of each variable measured as the mean decrease in the Gini index when the variable in question is included in a tree (for the exact values please refer to Supplement section~\ref{supp:randomforest}). 
\begin{figure}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.7\textwidth]{figure/unnamed-chunk-1-1}

@@

%The rows in the table are ordered according to descending importance. 
The variables with the most predictive power are cross-correlation and the overall number of matches, followed by the total depth of joint striations $S$ and total number of non-matches. CMS is found only in sixth place.

Besides including results from known matches against known non-matches, we can increase the number of comparisons in the Hamby study to include all possible land-to-land comparisons. This effectively doubles the number of data points available. Comparisons not previously included in fitting the random forest can also be used as an additional source for assessing error rates.  Results for this and a more detailed discussion can be found in Supplement section~\ref{supp:extended}.

\section{Conclusion}

We have presented an algorithm which detects the most prominent but least relevant structure of a bullet from a firearms identification perspective, removes these features, and produces residuals which allow for the easy identification of markings. We have generalized this algorithm to align the residuals from two bullets to automatically determine whether they are matches. We created a random forest model using many of the features to provide a probabilistic assessment of the quality of a match, along with the most relevant features. Matching bullets is clearly not a single step, but rather a sequence of a number of data analysis tasks which each deserve attention. As there is no scientific standard in place at this point in time, our intent is to explain one approach to addressing these tasks, while documenting all steps and providing all code so other researchers and forensic scientists can reproduce and expand on our findings.

The matching algorithm is sensitive to the parameter choices made. The heights at which signatures are extracted (currently 25$\mu m$ apart) to evaluate stability, as well as the cross-correlation factor (currently 0.95) we set as a minimum threshold do affect the final outcome. Another parameter that must be selected is the amount of smoothing when identifying peaks and grooves (currently, a window of 23.4375$\mu m$ is used, corresponding to a window of 7 values to the left and the right of an observation). While we tried to lay out in the paper the impact that each of the parameter choices has on the matching performance, we are still far from an optimized scenario.

The Hamby study serves as our evaluation `database'. It consists of only  35 bullets -- this is obviously not a particularly realistic scenario for an automatic matching procedure, but for now we are unaware of other databases containing bullets in the x3p format.

>>>>

Each fired bullet is provided in the form of a set of six x3p files, where each file is a surface scan between adjacent grooves on the bullet, called a ``land". In the Hamby data, typical length (groove-to-groove) of a land is about 1998.28 micrometers or 2 millimeters. For notational simplicity, we refer to a particular land of a bullet as bullet X-Y, where X is the bullet identifier, and Y is the land number. An example of plotting one of these lands is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}. These figures show side and top profiles of the land respectively. The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artifact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clockwise or counter-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic, i.e.\ a feature that pertains to a particular class of firearms, and is not unique at the individual barrel/bullet level.

@@

\end{knitrout}
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between signatures (c).}
\end{figure}
%\afterpage{\clearpage}
Figure~\ref{fig:match} gives an overview of the automated matching routine: 
We first identify a stable region for each bullet land and extract the signature at the lowest height in this region, because typically, individual characteristics are best expressed at the lower end of the bullet (see Supplement Section~\ref{supp:bulletbottom} for a more detailed discussion).

All of the other steps are done on pairs of bullet lands:
\begin{enumerate}
\item {\bf Smooth the two signatures} using a loess with a very small span (see Figure~\ref{fig:smooth}). 
\item Use cross-correlation to {\bf find the best alignment} of the two signatures: shift one of the signatures by the lag indicated by the cross-correlation function (see Figure~\ref{fig:ccf} for the cross-correlation function and Figure~\ref{fig:crosscutX} for the resulting shift).
\item Using a rolling average, {\bf identify peaks and valleys} for each of the signatures. We then define an interval around the location of the extrema on each side as one third of the distance to the location of the next extrema (see Figure~\ref{fig:smoothcutb}). Peaks and valleys constitute the \emph{striation marks} on the bullet.
\item {\bf Match striations across signatures:} based on the intervals around the extrema as defined above, we identify common intervals as the areas in which two or more of the individual intervals overlap: a joint interval is defined as the smallest interval that encompasses all of the overlapping intervals. A joint interval is then called a match(ing stria) between the signatures, if all of the intervals are of the same type of extrema, i.e.\ they are either all peaks or all valleys. In Figure~\ref{fig:match} all matches are shown as color-filled rectangles corresponding to their type of extrema (peaks are shown in orange, and valleys in green). Non-matching intervals are left grey. 
\item {\bf Extract features from the aligned signatures and the matches between them:} many different features can be extracted from the aligned signatures. Here, we describe a few of the ones that can be found in the literature and some that we found to be of practical relevance:
\begin{enumerate}[label=(\roman*)]
\item Maximal number of CMS (consecutive matching striae), and, similarly, the number of consecutively non-matching striae (CNMS), 
\item Number of matches and non-matches,
\item The value of the cross-correlation function (ccf) between the aligned signatures \citep{vorburger:2011},
\item Average difference $D$ between signatures, defined as the Euclidean vertical distance between surface measurements of aligned signatures. Let $f(t)$ and $g(t)$ be smoothed, aligned signatures:
\[
D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2,
\]
\item The sum $S$ of average absolute heights of matched extrema: for each of the two matched stria, compute the average of the absolute heights of the peaks or valleys. $S$ is then defined as the sum of all these averages. 
\end{enumerate}
\end{enumerate}
%
The difference $D$ between signatures is here defined as the Euclidean distance (in $\mu m$). In the paper by \citet{ma:2004}, distance is defined as a measure relative to the first signature, which serves as a comparison reference and is therefore a unitless quantity.

Counting the maximal number of CMS is part of the current practice to identify bullet matches~\citep{nichols:1997, nichols:2003, nichols:2003b}. 
In the example of Figure~\ref{fig:match}, the number of consecutive matching striations (CMS) is fifteen, a high number suggestive of a match between the lands. Note that the definition of CMS we use does not match the one given in \citet{thompson:2013}. There, CMS is defined only in terms of matching peaks without regarding valleys. Additionally, peaks in  \citet{thompson:2013} are  used only if they can be identified and matched `within a tolerable range' between lands. The definition given here is computationally less complex, but should yield highly correlated values, because of the requirement to only consider signatures from a stable region in the land (see Section~\ref{sec:heights} for further details on stability of regions). In the Hamby study, the definition of CMS by \citet{thompson:2013} leads to approximately half of the values of CMS defined in this paper (with a correlation coefficient between the values of the two definitions of about 0.92). 
For lead bullets, such as used in the Hamby study, \citet{biasotti:1959} considered four or more consecutive peaks (corresponding to eight or more consecutive lines in our definition) to be sufficient evidence of a match.  %We will assess the claim that eight or more consecutive lines are sufficient for a match in the next section.

@@

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities between signatures. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/crosscuts-1}

@@

\end{knitrout}
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
For this alignment we use the cross-correlation function to find a maximal amount of agreement between the signatures \citep{bachrach:2002, chu:2010, vorburger:2011, thompson:2013}.
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

@@

\section{Evaluation}
In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance using the James Hamby study data. As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we compute the cross-correlation of signatures extracted from heights 25$\mu m$ apart. For a stable region, we require a minimum of 0.95 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands has scratch marks across the surface, also known as `tank rash'.
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read_x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 9 Bullet 2-4}
\includegraphics[width=\textwidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet B-2}
\includegraphics[width=\textwidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet Q-4}
\includegraphics[width=\textwidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}Images of the four lands that got flagged during the quality assessment. All of them show scratch marks (tank rash) across the striation marks from the barrel. They are excluded from the remainder of the analysis.}
\end{figure}
%
We exclude these four lands from further matching considerations and
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are comparing $15 \times 6 -2 = 90 - 2 = 88$ lands from unknown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). 
Ideally, results look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land C-3 and all 118 lands from known bullets. Two lands show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and~\ref{fig:hamby-perfect}c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all lands, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

@@

For smaller numbers of CMS, the percentage of false positives quickly increases. However, if we take other features of the image into account, we can increase the number of correct matches considerably: Figure~\ref{fig:densities} gives an overview of the densities of all of the features derived earlier, for known matches (KM) and known non-matches (KNM). The densities of almost all of the features show strong differences between matches and non matches. For example, a high amount of cross-correlation between two signatures is indicative of a match --  in the Hamby study, only known matches have a cross-correlation of 0.75 or higher. There are 97 land-to-land comparisons with a cross-correlation that high.

@@

%Similarly, the Euclidean distance $D$ between two signatures is strongly correlated to known matches. Out of the 48 pairs of land-to-land signatures with a distance of less than .25, 47 correspond to known matches (the single known mismatch consists of a pair of signatures that both are very flat with peaks and valleys of less than 1.5$\mu m$). %The average distance between known matches in the Hamby study is $mean(bstats$D[bstats$match])\mu m$ with a standard deviation of $sd(bstats$D[bstats$match])\mu m$; while known mismatches have an average  distance of  $mean(bstats$D[!bstats$match])\mu m$ with a standard deviation of $sd(bstats$D[!bstats$match])\mu m$.
All of the features in Figure~\ref{fig:densities} show large, if not significant, differences between matches and non-matches. The predictive power of each one of these features is shown in the form of the Receiving Operating Characteristic (ROC) curves in Figure~\ref{fig:rocs}. The features are arranged in descending order according to the area under the curve (AUC).
The dots mark the equal error rate, i.e. the location on the ROC curve, where false positive and false negative error rates are the same. The smaller the value, the better. We see that in this instance a low equal error rate (EER) goes hand in hand with high predictive power as measured in AUC.
The feature with the highest individual predictive power is $S$, the sum of the average heights of two signatures at peaks and valleys. The maximal number of CMS is only in the seventh position here. The overall high AUC values indicate that we can successfully employ machine learning methods to distinguish matches from non-matches.
%Table~\ref{tab:importance} shows an overview in the averages and corresponding standard deviations of the set of feature variables derived in section~\ref{sec:algorithm} from the 3D topological surface measurements. There are quite large differences in these averages, indicating

@@

\end{knitrout}
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 100 trees, the error rate of land-to-land comparisons stabilizes at 0.0039. This is a weighted average between false positive error rate of 0.0001 and an error rate of false negatives of 0.2267. This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known  matches and non-matches (see Figure~\ref{fig:tree-forest}).
Note that this error rate is based on land-to-land comparisons and is expected to be much lower for bullet-to-bullet comparisons. In the case of the Hamby data, even a single tree results in an overall error rate of zero, if we require that a match of two bullets occurs when at least two of the bullet's lands are matched. % (see Figure~\ref{fig:treeresult}). 
%' \begin{figure}[hbtp]
%'   \centering
%' <<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\textwidth'>>=
%' bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
%' bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
%'   n = n(),
%'   pred = sum(pred > 0.5),
%'   forest = sum(forest > 0.5)
%' )
%' library(reshape2)
%' bm <- melt(data=bullets, id.var=c("bullet", "n"))
%' 
%' bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
%' qplot(data=bm, 100*value/n, reorder(bullet, value/n), size=I(2.5), 
%'       shape=prediction, colour=prediction, alpha=I(0.6)) + 
%'   theme_bw()  + ylab("") + 
%' #  scale_shape_manual(values=c(1,2)) +
%'   scale_x_continuous("Percent of correctly predicted land-to-land matches", 
%'                      breaks = c(0,25,50,75,100), limits=c(0,100)) +
%'   theme(legend.position = "bottom")
%' @
%' \caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. In the tree, at least \hh{40\% of all possible} lands of a bullet are  matched correctly, while the random forest achieves perfect results. Unknown bullet B is the hardest for the tree to match.}
%' \end{figure}
%' %
This makes the errors in the automated approach smaller than the human error in the Hamby study. Out of the 507 participants who returned results, eight (out of $15 \times 507 = 7,605$) bullets were not matched conclusively, corresponding to a rate of 0.0011. 
%\hh{XXX calling this a false negative rate is problematic. FEs will be up in arms, because inconclusive doesn't necessarily mean something was missed, it could also mean that there was something wrong with the bullets. XXX }

For the Hamby data, error rates based on bullet-to-bullet matches do not carry a lot of weight because of the small size of the study: fifteen unknown bullets are successfully matched to two pairs of ten bullets. Matching bullets can only be tested realistically in a much bigger experiment. 
Another thing to note about  the random forest's error rates is that they are based on probability cutoffs of 0.5, i.e.\ whenever the predicted probability of a match exceeds 0.5, a match is declared. Basing this decision on a threshold fixed at 0.5 may not be the best approach. In practice, examiners are allowed a third option of `inconclusive'. On a probability spectrum of outcomes we could therefore introduce an interval of `inconclusive' results in the middle of the spectrum -- which turns out to be unnecessary in the Hamby study, because, here, the results from the random forest are very clear cut. Figure~\ref{fig:tree-forest} shows a comparison of the predicted probabilities of a match by the tree and the random forest. As expected, the random forest provides a more realistic estimate of the uncertainty in the classification. %\hh{The tree predicts 132 known matches correctly, with 40 false negatives and one false positive match. The random forest predicts all 172 known matches correctly as well as all 10,212 known non-matches.} 
%Overall, there is a lot of agreement; out of 10,800 land-to-land comparisons, only four cases result in a different decision.

@@

Besides resulting in a probabilistic quantification of matches, random forests also provide an assessment of the importance of each of the features derived from the bullets' 3D topological surface measurements. Figure~\ref{fig:importance} shows an overview of the importance of each variable measured as the mean decrease in the Gini index when the variable in question is included in a tree (for the exact values please refer to Supplement Section~\ref{supp:randomforest}). 
\begin{figure}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.7\textwidth]{figure/unnamed-chunk-1-1}

@@

%The rows in the table are ordered according to descending importance. 
The variables with the most predictive power are cross-correlation and the overall number of matching extrema, followed by the total depth of joint striations $S$ and total number of non-matches. CMS is found only in sixth place.

Besides including results from known matches against known non-matches, we can increase the number of comparisons in the Hamby study to include all possible land-to-land comparisons. This effectively doubles the number of data points available. Comparisons not previously included in fitting the random forest can also be used as an additional source for assessing error rates.  Results for this and a more detailed discussion can be found in Supplement Section~\ref{supp:extended}.

\section{Discussion}

We present an algorithm which detects the most prominent but least relevant structure of a bullet from a firearms identification perspective, removes these features, and produces residuals which allow for the easy identification of markings. We have generalized this algorithm to align the residuals from two bullets to automatically determine whether they are indistinguishable. A random forest model provides a probabilistic assessment of the strength of a match, along with an ordering of the relevance of features. Matching bullets is clearly not a one-step process, but rather a sequence of data analysis tasks each deserving attention. As there is no scientific standard in place at this point in time, our intent is to explain an approach to addressing these tasks, while documenting all steps and providing all code so other researchers and forensic scientists can reproduce and expand on our findings.

The matching algorithm is sensitive to the parameter choices made. The heights at which signatures are extracted (currently 25$\mu m$ apart) to evaluate stability, as well as the cross-correlation factor (currently 0.95) we set as a minimum threshold do affect the final outcome. Another parameter that must be selected is the amount of smoothing when identifying peaks and grooves (currently, a window of 23.4375$\mu m$ is used, corresponding to a window of 7 values to the left and the right of an observation). We try to lay out in the paper the impact that each of the parameter choices has on the matching performance, but more research and better data are needed to define an optimized scenario.

The Hamby study serves as our evaluation `database'. It consists of only  35 bullets -- this is obviously not a particularly realistic scenario for an automatic matching procedure, but for now we are unaware of other databases containing bullets in the x3p format that we could add to our study.
